Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-21 ms.tex
UL changes

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with  elements $w_k$ such that $w_k \ne w_j$ for some $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type; 
$C$ is a positive diagonal  matrix with  elements $C_k$;
$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^n)_{j,k} > 0$ for all $j,k$ for some natural $n$.
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness.

@@

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation $f'$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= \big(1 - \delta C_0\big) \frac{w_0}{\bar{w}} f_0 + \beta C_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
f_n' &= \big(1 - \beta C_n\big) \frac{w_n}{\bar{w}} f_n + \delta C_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{w_k}{\bar{w}} f_k + \\
	 & \delta C_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta C_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$.
Here $\bar{w}=\sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.

@@

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.

Following selection and migration the new frequencies $f'_k$ are given exactly by Eq.~\ref{eq:model_sum}.
If columns of ${M}$ are identical
\begin{equation}
{M} = \begin{bmatrix}
m_1 & m_1 & \ldots & m_1 \\
m_2 & m_2 & \ldots & m_2 \\
\vdots & \vdots & \vdots & \vdots \\
m_n & m_n & \ldots & m_n
\end{bmatrix},
\end{equation}
with $m_k>0$ and $\sum_{k=1}^{n}{m_k}=1$,
then $m_k$ can be considered the relative population size of deme $A_k$ --
this is the non-homogeneous extension of Deakin's \emph{homing model}~\citep{Deakin1966,Karlin1982}.

@@

\noindent{\bf Result 1. Mean fitness principle.}
{\sl Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors.
Then,
\begin{equation}\label{mfp}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

@@

\begin{proof}
$\hat{\bar w}$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{\hat v_k} = 1$ and $\sum_k{\hat u_k \hat v_k} = 1$.

Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v.
\end{equation}

@@

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
\hat u\tr \hat{\bar w} = \hat u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\hat{\bar w}$ and the $k$
element of $u$:
\begin{equation}
\hat u_k \hat{\bar w} = (1-C_k) w_k \hat u_k + C_k w_k \hat u\tr [M]_k.
\end{equation}
Multiplying both sides by $\hat v_k$  and rearranging, we get 
\begin{equation}
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k) = \hat v_k w_k (\hat u\tr [M]_k - \hat u_k),
\end{equation}
which when substituted into Eq.~\ref{mfp_intermediate} yields:
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k).
\end{equation}
Finally, note that $\hat u_k, \hat v_k, C_k \ge 0$ and therefore
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat{\bar w} - w_k),
\end{equation}
which completes the proof.
\end{proof}

@@

Given $0 < w_k\le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.

@@

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$ in the space $\sum_{k=1}^n{\hat v_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

@@

Note that if the modifier has the same effect on all individuals (i.e., $C_k = \hat\alpha$ and $\tilde{C}_k = \alpha$ for all $k$) then $\cl_{ex}$ becomes
\begin{equation}
\cl_{ex} = ((1 - \hat\alpha)I + \hat\alpha M) D /\hat{\bar w},
\end{equation}  
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha < \hat\alpha$, will be favored by natural selection.
Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above, we find
$$
\frac{\partial \hat{\bar w}}{\partial \alpha} = 
\hat{\bar w} - 1 \le 0,
$$
because $\hat{\bar w} = \sum_{k=1}^{n}{\hat v_k w_k}$, and $w_k \le 1$ for all $k$.

@@

Now, let $\norm{x} = \sum_{k=1}^{n}\norm{x_k}$, so that $x_t \goesto x$ implies
\begin{equation}
\norm{x_t} \goesto \norm{x},
\label{eq:norm_limit}
\end{equation}
as $t \goesto \infty$ and rewrite Eq.~\ref{eq:model} as 
\begin{equation}
f' = \frac{A f}{\norm{A f}}. 
\label{eq:model_norm}
\end{equation}

It is easily seen that $f^t =A^tf/\|A^tf\|$. Hence 
\begin{multline}
\frac{A^t f}{\norm{A^t f}} = 
\frac{\frac{1}{(\hat{\bar w})^t} A^t f}{\frac{1}{(\hat{\bar w})^t} \norm{A^t f}} = 
\frac{\big(\frac{A}{\hat{\bar w}}\big)^t \cdot f }{ \norm{ \big(\frac{A}{\hat{\bar w}}\big)^t \cdot f }} \goesto
\frac{\alpha \hat v}{\norm{\alpha \hat v}} = \hat v,
\end{multline}
as $t \goesto \infty$, since $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
This completes the proof.
\end{proof}

>>>>

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with  elements $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
%and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type; 
$C$ is a positive diagonal  matrix with  elements $C_k$;
$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^n)_{j,k} > 0$ for all $j,k$ for some natural $n$.
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness $\bar w=\sum_{k=1}^n f_kw_k$.

@@

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation $f'_k$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= \big(1 - \delta C_0\big) \frac{w_0}{\bar{w}} f_0 + \beta C_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
f_n' &= \big(1 - \beta C_n\big) \frac{w_n}{\bar{w}} f_n + \delta C_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{w_k}{\bar{w}} f_k + \\
	 & \delta C_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta C_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$.
Here $\bar{w}=\sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.

@@

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where the matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.

Following selection and migration the new frequencies $f'_k$ are given exactly by Eq.~\ref{eq:model_sum}.
If the columns of ${M}$ are identical
\begin{equation}
{M} = \begin{bmatrix}
m_1 & m_1 & \ldots & m_1 \\
m_2 & m_2 & \ldots & m_2 \\
\vdots & \vdots & \vdots & \vdots \\
m_n & m_n & \ldots & m_n
\end{bmatrix},
\end{equation}
with $m_k>0$ and $\sum_{k=1}^{n}{m_k}=1$,
then $m_k$ can be considered the relative population size of deme $A_k$ --
this is the non-homogeneous extension of Deakin's \emph{homing model}~\citep{Deakin1966,Karlin1982}.

@@

\noindent{\bf Result 1. Mean fitness principle.}
{\sl Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors, such that $\sum_k\hat v_k=1$ and $\sum_k\hat u_k\hat v_k=1$..
Then,
\begin{equation}\label{mfp}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

@@

\begin{proof}
%$\hat{\bar w}$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
%and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{\hat v_k} = 1$ and $\sum_k{\hat u_k \hat v_k} = 1$.
Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v.
\end{equation}

@@

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
\hat u\tr \hat{\bar w} = \hat u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\hat{\bar w}$ and the $k$
element of $u$:
\begin{equation}
\hat u_k \hat{\bar w} = (1-C_k) w_k \hat u_k + C_k w_k \hat u\tr [M]_k.
\end{equation}
Multiplying both sides by $\hat v_k$  and rearranging, we get 
\begin{equation}
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k) = \hat v_k w_k (\hat u\tr [M]_k - \hat u_k),
\end{equation}
which when substituted into Eq.~\ref{mfp_intermediate} yields:
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k).
\end{equation}
Finally, note that $\hat u_k, \hat v_k, C_k > 0$ and therefore
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat{\bar w} - w_k),
\end{equation}
which completes the proof.
\end{proof}

@@

Given $w_k>0$, from Eq.~\ref{mfp_intermediate} we have the following corollary.

@@

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(\hat v,0)$ in the space $\sum_{k=1}^n{v_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

@@

Note that if the modifier has the same effect on all individuals (i.e., $C_k = \hat\alpha$ and $\tilde{C}_k = \alpha$ for all $k$) then $\cl_{ex}$ becomes
\begin{equation}
\cl_{ex} = ((1 - \alpha)I + \alpha M) D /\hat{\bar w},
\end{equation}  
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha < \hat\alpha$, will be favored by natural selection.
Setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in Eq.\ 25, we find
$$
\frac{\partial \bar w}{\partial \alpha} = 
\frac{1}{\alpha}\left[\bar w-\sum_{k=1}^nu_kv_kw_k\right]
$$
(following~\citet{Altenberg2017}, $\frac{\partial\bar w}{\partial \alpha}<0$).

@@

Now, let $\norm{x} = \sum_{k=1}^{n}|x_k|$, so that $x_t \goesto x$ implies
\begin{equation}
\norm{x_t} \goesto \norm{x},
\label{eq:norm_limit}
\end{equation}
as $t \goesto \infty$ and rewrite Eq.~\ref{eq:model} as 
\begin{equation}
f' = \frac{A f}{\norm{A f}}. 
\label{eq:model_norm}
\end{equation}

It is easily seen that $f^t =A^tf/\|A^tf\|$. Hence 
\begin{multline}
\frac{A^t f}{\norm{A^t f}} = 
\frac{\frac{1}{(\hat{\bar w})^t} A^t f}{\frac{1}{(\hat{\bar w})^t} \norm{A^t f}} = 
\frac{\big(\frac{A}{\hat{\bar w}}\big)^t \cdot f }{ \norm{ \big(\frac{A}{\hat{\bar w}}\big)^t \cdot f }} \goesto
\frac{\alpha \hat v}{\norm{\alpha \hat v}} = \hat v,
\end{multline}
since $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
This completes the proof.
\end{proof}
